---
title: "Date-Control Regression Results"
author: "Owen Melia"
date: "11/11/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Load Data

```{r load data}
data_dir <- "~/projects/metabolomics/data/date_controlled/"


resid_df <- read.csv(file.path(data_dir, 'imputed_data_date_controlled_resids.txt'))
coeff_df <- read.csv(file.path(data_dir, 'imputed_data_date_controlled_coefficients.txt'))
pval_df <- read.csv(file.path(data_dir, 'imputed_data_date_controlled_pvals.txt'))
```

## Look at Residuals

```{r load_residuals}
resid_df <- read.csv('~/projects/metabolomics/data/date_controlled/imputed_data_date_controlled_resids.txt')
rownames(resid_df) <- resid_df$X
resid_df$X <- NULL
```


```{r residual normality test}
testnormality = function(df, print_bad_cols = FALSE, threshold = 0.01)
{
  cc=0
  pvec = rep(NA,ncol(df))
  for(ii in 1:ncol(df)){
    pp = shapiro.test(df[,ii])$p
    pvec[ii] = pp
    if(pp < threshold){
      if(print_bad_cols){
        print("-----")
        print(colnames(df)[ii])
        print(pp)
      }
      cc=cc+1
    }
  }
  print(paste0("Non-normal columns according to Shapiro < ", threshold))
  print(cc)
  pvec
}

resid_df_normality_pvec <- testnormality(resid_df)

```

## Residuals From Inverse-Normalized Regression

I ran the same regression but with the columns of the metabolomics data matrix inverse-normalized. I'll load the data and run the Shapiro test.

```{r inverse_norm regression}
resid_inv_norm_df <- read.csv(file.path(data_dir, 'imputed_data_date_controlled_inv_normalized_resids.txt'))
rownames(resid_inv_norm_df) <- resid_inv_norm_df$X
resid_inv_norm_df$X <- NULL
inv_norm_pvec <- testnormality(resid_inv_norm_df)
```


When viewing the next plot, I had to remind myself that smaller p-values indicate that the sample is not normally distributed. The x axis plots p-values on the log scale, so the smaller p-values are on the left. 

A vertical line at the $\alpha = 0.01$ threshold is drawn
```{r shapiro_test_pvals}

shapiro_pval_df <- data.frame(standard = (resid_df_normality_pvec), 
                      inverse_normalized = (inv_norm_pvec))

shapiro_pval_df <- shapiro_pval_df %>% pivot_longer(everything(), names_to = 'distribution', values_to = 'pval')
plt <- (ggplot(data = shapiro_pval_df, aes(pval, fill = distribution, color = distribution))
        + geom_density(alpha = 0.1)
        + labs(x = "Log of pvalue of Shapiro Test",
               y = "Density"
               )
        + geom_vline(xintercept = 0.01)
        + scale_x_log10()
        )
plt
```


## Look at PVals

```{r summarize pvals}

rownames(pval_df) <- pval_df$X
pval_df$X <- NULL

pval_df_t <-data.frame(t(pval_df))

summary(pval_df_t)

```


The pvalues are tricky to plot in a useful way. They are concentrated around zero, with a lot of outliers uniformely distributed between 0 and 1. 

```{r plot_pvals}

pval_df_t$metabolite <- rownames(pval_df_t)

pval_df_long <- pval_df_t %>% pivot_longer( -c(metabolite), names_to = 'Day', values_to = 'pval')

plt <- (ggplot(data = pval_df_long, aes(metabolite, Day, fill = pval))
        + geom_tile()
        + theme(axis.text.x = element_blank(),
                axis.ticks.x = element_blank(),
                axis.ticks.y = element_blank(),
                axis.text.y = element_blank())
        )
# plt

plt_2 <- (ggplot(data = pval_df_long, aes(pval))
          + geom_histogram(bins = 200))
# plt_2

```
































