---
title: "Examining the Effect of Covariates"
author: "Owen Melia"
date: "11/11/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)

source('~/projects/lab-tools/paper_themes.R')

source('~/projects/metabolomics/workflowr/code/load_imputed_data.R')
```

## Load Covariates

With the metabolite data came metadata about the experiment. I'd like to consider these as covariates and see if we can control for batch effects. 

```{r load_sample_metadata}

covariates <- read.csv('~/projects/metabolomics/data/metabolomics_converted/imputed-data-7-11-2016-sample-metadata.txt', sep = '\t')
summary(covariates)
```

It seems that we should consider all of these as factors

```{r load covariates as factors}
covariates <- read.csv('~/projects/metabolomics/data/metabolomics_converted/imputed-data-7-11-2016-sample-metadata.txt',
                       sep = '\t',
                       colClasses = c('factor'))
rownames(covariates) <- covariates$DIAG_code
n_unique <- function(x){length(levels(x))}
unique_factors <- sapply(covariates, n_unique)
unique_factors
```

## Investigating Redundant Covariates

I would guess that there is a lot of redundancy in these data. I will first only look at the batch IDs.

```{r drop_other_values}

list_to_drop <- c('DIAG_code', 
                  'PARAM_BOX_LOCATION', 
                  'PARAM_MANIFEST_ORDER', 
                  'PARAM_SPECIMEN_CONDITION',
                  'PARAM_LC_COLUMN')
for(i in list_to_drop){
  covariates[[i]] <- NULL
}

```

I will plot a correlation heatmap. The results shown here could be an artifact of the order that R assigned factor levels. 

```{r plot_corr_heatmap}
corr_heatmap <-function(df, title = "Correlation"){
  cor_df <- data.frame(cor(df))
  cor_df$rows <- rownames(cor_df)
  df_for_plot <- cor_df %>% pivot_longer(-rows, names_to = 'columns', values_to ='correlation')
  plt <- (ggplot(data = df_for_plot, aes(rows, columns, fill = correlation))
          + geom_tile()
          + theme(axis.text.x = element_text(angle = 90))
          + labs(title = title))
  
  return(plt)
}

cov_numeric <- data.frame(sapply(covariates, as.numeric))
bb <- corr_heatmap(cov_numeric)
  
bb
```

I think I will now only focus on `PARAM_RUN_DAY` and `BATCH_ID_LCNEG`.

```{r drop other batch IDs}
list_to_drop <- c('BATCH_ID_LCPOSLAT',
                  'BATCH_ID_LCPOSEAR',
                  'BATCH_ID_LCPOLAR')
for(i in list_to_drop){
  covariates[[i]] <- NULL
}
```

## Regressing Out Covariates

I will choose a random meatabolite to regress:

$$ Metabolite = \sum \beta_i \ \text{BATCH_ID_LCNEG}_i + \sum \beta_i \ \text{PARAM_RUN_DAY}_i $$

```{r run_one_regression}


covariate_regression <- function(response, covariates){
  covariates$metabolite <- response
  reg_obj  <- lm(metabolite ~ . , data = covariates)
  return(reg_obj)
}

single_reg_obj <- covariate_regression(imputed_df$X1.2.dilinoleoyl.GPC..18.2.18.2., covariates)

print(summary(single_reg_obj))
```


So it still looks like there are singularities. I will look at just the run day. 


## Regression on Run Day


```{r re construct covariates}


covariates <- read.csv('~/projects/metabolomics/data/metabolomics_converted/imputed-data-7-11-2016-sample-metadata.txt',
                       sep = '\t',
                       colClasses = c('factor'))
rownames(covariates) <- covariates$DIAG_code
list_to_drop <- c('DIAG_code', 
                  'PARAM_BOX_LOCATION', 
                  'PARAM_MANIFEST_ORDER', 
                  'PARAM_SPECIMEN_CONDITION',
                  'PARAM_LC_COLUMN',
                  'BATCH_ID_LCPOSLAT',
                  'BATCH_ID_LCPOSEAR',
                  'BATCH_ID_LCPOLAR',
                  'BATCH_ID_LCNEG')
for(i in list_to_drop){
  covariates[[i]] <- NULL
}
```

```{r new regression}
reg_obj <- covariate_regression(imputed_df$X1.7.dimethylurate, covariates)
print(summary(reg_obj))
```

## Look at Residuals

Here I am plotting the density of the residuals in red and the normal approximation in blue.
```{r plot_residuals}
resid_df <- data.frame(residuals = single_reg_obj$residuals)

plt <- (ggplot(data = resid_df, aes(residuals))
        + geom_density(color = 'red') 
        + bar_base_theme_()
        + stat_function(fun = dnorm, 
                        color = 'blue',
                        args = list(mean = mean(resid_df$residuals),
                                    sd = sd(resid_df$residuals))))

plt



```


The residuals for this one are not normal. I will regress out the day parameter using `data/regress_out_dates.R` and `data/regress_out_dates.sh` and I will continue the analysis in [Date-Control Regression Results](date_control_regression_results.html)


